# Assessment of the nature of batch effects {#simu}

## Simulations

### Mean = 5, and unequal variance

'Systematic' refers to homogeneous change amongst all microbial variables (OTUs) due to having the same source of variation. We illustrate this concept in a linear model framework where batch and treatment regression coefficients are estimated simultaneously on each OTU. For a given batch effect and a given OTU, we can formulate the systematic assumption as:

$$\beta_{j} \sim N(\mu,\sigma^{2})$$

where $\beta_{j}$ is the batch regression coefficient of OTU$_{j}$ ($j = 1,...,p$). Here we consider the simplest case of a linear model with one batch predictor, but this formulation could be extended to a model with multiple batch predictors where the batch regression coefficients can represent more than two batch levels. In a univariate model that tests each OTU individually, then the distribution of the batch coefficients of all OTUs is Gaussian with a mean $\mu$, and standard deviation $\sigma$. This indicates that the batch effect has a similar, though not necessarily identical, influence on all OTUs. 

To illustrate the type of batch effects, we simulated a set of data with 50 samples and 10,000 OTUs each, based on the simulation approach from [@gagnon2013removing].

The dataset with a systematic batch effect:

+ $\beta_{j} \sim N(5,1^{2})$ for $j=1,...,p$ OTUs;     
+ $\sigma_{j} \sim N(0,2^{2})$ for $j=1,...,p$ OTUs; This variance per OTU is aimed to simulate data as realistically as possible.       
+ $\beta_{ij} \sim N(\beta_{j}, \sigma_{j}^{2})$ for $i = 1,...,n$ samples.       


```{r}
# Create the simulated data
m <- 50
n <- 10000
nc <- 1000 # negative controls without treatment effects
p <- 1
k <- 1
ctl <- rep(FALSE, n)
ctl[1:nc] <- TRUE
# treatment effect
X <- matrix(c(rep(0, floor(m/2)), rep(1, ceiling(m/2))), m, p)
beta <- matrix(rnorm(p*n, 5, 1), p, n) #treatment coefficients
beta[ ,ctl] <- 0
# batch effect
W <- as.matrix(rep(0, m), m, k)
W[c(1:12,38:50), 1] <-  1
alpha <- matrix(rnorm(k*n, 5, 1), k, n)
Y_alpha <- sapply(alpha, function(alpha){rnorm(m, mean =  alpha, 
                                              abs(rnorm(1, mean = 0, sd = 2)))})
YY_alpha <- apply(Y_alpha, 2, function(x){x*W})

epsilon <- matrix(rnorm(m*n, 0, 1), m, n)
Y <- X%*%beta + YY_alpha + epsilon


# estimate batch coefficient for each OTU
w.cof <- c()
for(i in 1:ncol(Y)){
  res <- lm(Y[ ,i] ~ X + W)
  sum.res <- summary(res)
  w.cof[i] <- sum.res$coefficients[3,1]
}

par(mfrow = c(2,2))
hist(w.cof,col = 'gray')
plot(density(w.cof))
qqnorm(w.cof)
qqline(w.cof, col = 'red')
par(mfrow = c(1,1))
```

The histogram, density plot and quantile–quantile plot are plotted using estimated batch regression coefficients $\hat{\beta}_{j}$ for each OTU $j$. These plots display the distribution of batch coefficients is very close to normal, indicating a systematic batch effect.

### Mean = 0 or 5, and unequal variance

Non-systematic batch effects have a heterogeneous influence on microbial variables. Using a linear model framework, as described previously, we can formulate this non-systematic assumption as:

\[
\beta'_{j} \sim 
  \begin{cases} 
    N(0,\delta^{2}) & \text{for OTUs with no batch effect,} \\
    N(\mu,\sigma^{2}) & \text{for OTUs with batch effect.}
  \end{cases}
\]

Therefore, the batch regression coefficients $\beta'_{j}$ may follow skewed distributions with several modes.

The dataset with non-systematic batch effect:

+ $\beta'_{t} \sim N(0,1^{2})$ and $\beta'_{k} \sim N(5,1^{2})$ for $t=1,...,T$ OTUs, $k=1,...,K$ OTUs and $T=\frac{3}{4}p$, $K=\frac{1}{4}p$;     
+ $\sigma'_{j} \sim N(0,2^{2})$ for $j=1,...,p$ OTUs;       
+ $\beta'_{ij} \sim N(\beta'_{j}, \sigma_{j}^{'2})$ for $i = 1,...,n$ samples.       

```{r}
# Create the simulated data
m <- 50
n <- 10000
nc <- 1000 # negative controls without treatment effects
p <- 1
k <- 1
ctl <- rep(FALSE, n)
ctl[1:nc] <- TRUE
# treatment effect
X <- matrix(c(rep(0, floor(m/2)), rep(1, ceiling(m/2))), m, p)
beta <- matrix(rnorm(p*n, 5, 1), p, n) #treatment coefficients
beta[ ,ctl] <- 0
# batch effect
W <- as.matrix(rep(0, m), m, k)
W[c(1:12,38:50), 1] <-  1
alpha2 <- matrix(sample(c(rnorm(k*(3*n/4), 0, 1),rnorm(k*(n/4), 5, 1)), n), k, n)
Y_alpha2 <- sapply(alpha2, function(alpha){rnorm(m, mean =  alpha, 
                                                sd = abs(rnorm(1, mean = 0, sd = 2)))})
YY_alpha2 <- apply(Y_alpha2, 2, function(x){x*W})

epsilon <- matrix(rnorm(m*n, 0, 1), m, n)
Y2 <- X%*%beta + YY_alpha2 + epsilon




w.cof2 <- c()
for(i in 1:ncol(Y2)){
  res <- lm(Y2[ ,i] ~ X + W)
  sum.res <- summary(res)
  w.cof2[i] <- sum.res$coefficients[3,1]
}

par(mfrow = c(2,2))
hist(w.cof2, col = 'gray')
plot(density(w.cof2))
qqnorm(w.cof2)
qqline(w.cof2, col = 'red')
par(mfrow = c(1,1))
```

The histogram, density plot and quantile–quantile plot are plotted using estimated batch regression coefficients $\hat{\beta}_{j}$ for each OTU $j$, showing a bi-modal distribution. This distribution indicates a non-systematic batch effect. 

**We observed similar patterns in our real case studies, suggesting that the batch effects are mixed with multiple sources and are non-systematic (see the examples below).**

## Real data

We also estimate the batch regression coefficients for each OTU in real datasets to assess the nature of batch effects.

### Sponge data

```{r}
sponge.b.coeff <- c()
for(i in 1:ncol(sponge.tss.clr)){
  res <- lm(sponge.tss.clr[ ,i] ~ sponge.trt + sponge.batch)
  sum.res <- summary(res)
  sponge.b.coeff[i] <- sum.res$coefficients[3,1]
}

par(mfrow = c(2,2))
hist(sponge.b.coeff,col = 'gray')
plot(density(sponge.b.coeff))
qqnorm(sponge.b.coeff)
qqline(sponge.b.coeff, col='red')
par(mfrow = c(1,1))
```

In sponge data, the histogram, density plot and quantile–quantile plot are also plotted using estimated batch regression coefficients $\hat{\beta}_{j}$ for each OTU $j$. The bi-modal distribution indicates a non-systematic batch effect, which is similar as the results from simulation with non-systematic batch effects. 

### AD data

```{r}
ad.b.coeff <- c()
ad.batch.relevel <- relevel(ad.batch, '01/07/2016')
for(i in 1:ncol(ad.clr)){
  res <- lm(ad.clr[,i] ~ ad.trt + ad.batch.relevel)
  sum.res <- summary(res)
  ad.b.coeff[i] <- sum.res$coefficients[4,1]
}

par(mfrow = c(2,2))
hist(ad.b.coeff,col = 'gray')
plot(density(ad.b.coeff))
qqnorm(ad.b.coeff)
qqline(ad.b.coeff, col='red')
par(mfrow = c(1,1))
```
 
We observe similar results in AD data. The distribution of estimated batch regression coefficients is not normal. The batch effects therefore are non-systematic.